**The status of the microbial census: an update**

### 2015-11-13 Schloss_Census2_mBio_2015.Rmd

Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences. In the 13 years since the collection of data for Schloss and Handelsman’s analysis, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study. Although there has been a robust growth in the number of full-length sequences deposited to GenBank since its creation in 1983, the rate of growth has stalled over the past 5 years and the deposits have been dominated by a handful of research groups studying a limited number of environments. As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.

@@

* How have the number of full-length sequences changed over the past N years?
* How does this vary by environment?
* How has the coverage of the OTUs represented by these sequences changed over the last N years?
* How has the number of sequences per study changed over the past N years?
* How has Figure 1 changed over the past N years?

@@

@@

>>>>

Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences. In the 13 years since the collection of data for Schloss and Handelsman’s analysis, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study. Although there has been a robust growth in the number of full-length sequences deposited to GenBank since its creation in 1983, the rate of growth has stalled over the past 5 years and the deposits have been dominated by a handful of research groups studying a limited number of environments. As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments. In this, continued technological advances, such as the reconstruction of nearly full-length sequences from metagenomics with EMIRGE (Miller 2011) and inovations in culturing (references needed: iChip, Epstein dormancy, Thrash 2015; Gifford 2014), will be vital.

@@

* Figure 1A. How have the number of full-length sequences changed over the past N years?
* Figure 1B. How does this vary by environment?

* Figure 2. How has the number of sequences per study changed over the past N years?

* Figure 3. How has the coverage of the OTUs represented by these sequences changed over the last N years?

* How has Figure 1 changed over the past N years?

* In what environments do we have the best handle on extant diversity?
  Figure 4A,B. What are the differences in sampling effort/sequence coverage at the 97% level for the different coarse habitats, and how is this different between Bacteria and Archaea?
  Figure 5A,B. For the most sampled coarse habitats (Zoonotic for Bacteria, Aquatic for Archaea), what are the fine-scale habitats where they have been sampled most heavily?

* In what habitats do we have the poorest understanding of diversity (and thus expect to yield considerable new knowledge over the years)?
* How is this different for Bacteria vs. Archaea?
* What are the most significant improvements since 2004, and where are we still lacking the most data?
* For which phyla has there been substantial increases in sampled diversity since 2004?
* Can we make any recommendations for habitats where biodiversity is under-sampled, perhaps calls for conservation in certain environments based on how poorly we understand the microbial diversity?

@@

Impact of EMIRGE

Renewed call for cultivation

@@

**Figure 1. Sequences deposited by year**

Pat, this is a placeholder for you

**Figure 2. Sequence deposition by study**

Pat, ibid.

**Figure 3. Sampling by OTU cutoff**

I only have an old code from Rene for this, and I don't think it reports all the sequences

**Figure 4A. Bacterial sampling by environment**

```{r figure4a}
data <- read.table(file="data/mothur/all_bacteria.env_category.groups.rarefaction", header=T)
category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
plant_columns <- grep(category, colnames(data))
category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
aerosol_columns <- grep(category, colnames(data))
category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
aquatic_columns <- grep(category, colnames(data))
category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
built_columns <- grep(category, colnames(data))
category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
other_columns <- grep(category, colnames(data))
category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
soil_columns <- grep(category, colnames(data))
category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
zoonotic_columns <- grep(category, colnames(data))
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per.pdf")

par(mar=c(5,5,0.5,0.5))
plot(NA, type="n", ylim=c(0,max(40000)),
     xlim=c(0,max(data$numsampled)),
     xlab="Number of sequences sampled",
     ylab="",
     axes=FALSE)

points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
points(data$X0.03.aerosol~data$numsampled, type="l", col="green", lwd=2)
points(data$X0.03.aquatic~data$numsampled, type="l", col="blue", lwd=2)
points(data$X0.03.built~data$numsampled, type="l", col="orange", lwd=2)
points(data$X0.03.other~data$numsampled, type="l", col="black", lwd=2)
points(data$X0.03.soil~data$numsampled, type="l", col="purple", lwd=2)
points(data$X0.03.zoonotic~data$numsampled, type="l", col="deeppink", lwd=2)

legend(x=5e5, y=20000, lwd=2, lty=1,
       legend=c("Other", "Plant", "Aerosol", "Aquatic", "Built", "Soil", "Zoonotic"),
       col=c("black", "red", "green", "blue", "orange", "purple", "deeppink"))

axis(1, at=seq(0,700000,1e5), label=c("0", "100,000", "200,000", "300,000", "400,000", "500,000", "600,000", "700,000"), cex.axis=0.8)
axis(2, at=seq(0,40000,5000), label=c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "35,000", "40,000"), las=2, cex.axis=0.8)
box()
mtext(side=2, "Number of OTUs", line=4)

dev.off()

**Figure 4B. Archaeal sampling by environment**

```{r figure4b}

data <- read.table(file="data/mothur/all_archaea.env_category.groups.rarefaction", header=T)
category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
plant_columns <- grep(category, colnames(data))
category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
aerosol_columns <- grep(category, colnames(data))
category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
aquatic_columns <- grep(category, colnames(data))
category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
built_columns <- grep(category, colnames(data))
category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
other_columns <- grep(category, colnames(data))
category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
soil_columns <- grep(category, colnames(data))
category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
zoonotic_columns <- grep(category, colnames(data))
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
pdf(file="/Users/renegirard7gfy/Desktop/Arc_Uniq_3per.pdf")

par(mar=c(5,5,0.5,0.5))
plot(NA, type="n", ylim=c(0,max(data[,-1], na.rm=T)),
     xlim=c(0,max(data$numsampled)),
     xlab="Number of sequences sampled",
     ylab="",
     axes=FALSE)

points(data$X0.03.plant~data$numsampled, type="l", col="red", lwd=2)
points(data$X0.03.aerosol~data$numsampled, type="l", col="green", lwd=2)
points(data$X0.03.aquatic~data$numsampled, type="l", col="blue", lwd=2)
points(data$X0.03.built~data$numsampled, type="l", col="orange", lwd=2)
points(data$X0.03.other~data$numsampled, type="l", col="black", lwd=2)
points(data$X0.03.soil~data$numsampled, type="l", col="purple", lwd=2)
points(data$X0.03.zoonotic~data$numsampled, type="l", col="deeppink", lwd=2)

legend(x=15000, y=1550, lwd=2, lty=1,
       legend=c("Other", "Plant", "Aerosol", "Aquatic", "Built", "Soil", "Zoonotic"),
       col=c("black", "red", "green", "blue", "orange", "purple", "deeppink"))

axis(1, at=seq(0,35000,5000), label=c("0", "5,000", "10,000", "15,000", "20,000", "25,000", "30,000", "35,000"), cex.axis=0.8)
axis(2, at=seq(0,5250,750), label=c("0", "750", "1500", "2250", "3,000", "3750", "4500", "5250"), las=2, cex.axis=0.8)
box()
mtext(side=2, "Number of OTUs", line=4)

dev.off()
```
